---
title: "Baseball Analytics"
subtitle: "A comprehensive analysis of factors that influence at-bat outcomes"
format:
  revealjs: 
    # embed-resources: true
    slide-number: true
    chalkboard: 
      buttons: false
    preview-links: auto
    css: styles.css
    footer: STAT 405/605 - Andersen Pickard, Aly Khanmohamed, Jacob Andreini, Enyu Li
server: shiny
---

# Introduction  {background="#6B6B6B"}

## Background {background="#a6bee3"}
- We wondered if various factors, such as at-bat length and pitcher/batter handedness, impacted at-bat outcomes
- We were curious if longer at-bats favored the hitter or pitcher more. In other words, who wears down faster?
- Additionally, we wondered if our data could point to a potential bias in end-of-season awards voting

## Data Sources {background="#a6bee3"}
- Main dataset from Retrosheet: 1,228,237 rows and 128 columns
  - Contained play-by-play data from MLB regular season and postseason, 2018-24

- Secondary dataset from Baseball Savant: 617 rows and 15 columns
  - Contained individual player data from 2024 regular season, min. 10 plate appearances

# At-Bat Length Data   {background="#6B6B6B"}

## Pitch vs. Foul Ball Totals  {background="#a6bee3"}

- Pitch totals correlates to foul ball totals
- As pitches increase, foul balls also increase

::: {.center}
```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 4.5
library(DBI)
library(RSQLite)

library(DBI)
library(RSQLite)

# Connect to the SQLite database
dcon <- dbConnect(RSQLite::SQLite(), "405baseball.sqlite")

# Load the CSV file (replace "path_to_your_data.csv" with the actual path to your file)
data <- read.csv("stat405baseball.csv")

# Write the data to the SQLite database, creating the stat405baseball table
dbWriteTable(dcon, "stat405baseball", data, overwrite = TRUE)

# Check if the table is now in the database
tables <- dbListTables(dcon)

library(dplyr)
library(ggplot2)
library(viridis)
library(showtext)
library(tidyr)
library(stringr)
font_add_google("Montserrat", "montserrat")
showtext_auto()
dcon <- dbConnect(RSQLite::SQLite(), "405baseball.sqlite")
query1 <- "SELECT 
  f.batter,
  f.total_fouls,
  f.plate_appearances AS pa_foul,
  f.total_fouls * 1.0 / f.plate_appearances AS fouls_per_pa,
  n.total_nump,
  n.plate_appearances AS pa_nump,
  n.total_nump * 1.0 / n.plate_appearances AS nump_per_pa
FROM
  (SELECT 
     batter,
     SUM(LENGTH(pitches) - LENGTH(REPLACE(pitches, 'F', ''))) AS total_fouls,
     COUNT(*) AS plate_appearances
   FROM stat405baseball
   GROUP BY batter) f
LEFT JOIN
  (SELECT 
     batter,
     SUM(nump) AS total_nump,
     COUNT(*) AS plate_appearances
   FROM stat405baseball
   GROUP BY batter) n
ON f.batter = n.batter"
result <- dbGetQuery(dcon, query1)
lm_model <- lm(total_nump ~ total_fouls, data = result)
par(bg = "#f9f9f9", mar = c(5, 5, 4, 2))
plot(
  result$total_fouls, result$total_nump,
  xlab = "Foul Balls Taken",
  ylab = "Pitches Seen",
  main = "Fouls vs. Pitches (2018–2024)",
  ylim = c(0, 20000),
  col = "#2c3e50",        
  pch = 19,              
  cex = 0.5,               
  cex.lab = 1.2,           
  cex.main = 1.4,          
  col.lab = "#34495e",     
  col.main = "#2c3e50",    
  bty = "n"               
)
grid(col = "#dddddd", lty = "dotted")
abline(lm_model, col = "#2980b9", lwd = 2.5) 
legend("topleft", legend = "Linear Fit", col = "#2980b9", lwd = 2.5, bty = "n")
```
:::

## Pitches per Plate Appearance  {background="#a6bee3"}

- Most common number of pitches: 4
- Longer at-bats become increasingly more rare

::: {.center}
```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 4.5
baseballdata <- dbGetQuery(dcon, "SELECT * FROM stat405baseball")
par(bg = "#f9f9f9", mar = c(5, 5, 4, 2))  
h <- hist(
  baseballdata$nump,
  breaks = 20,
  col = "#74b9ff",         
  border = "white",        
  ylim = c(0, 230000),
  xlim = c(0, 12),
  xlab = "Number of Pitches in Plate Appearance",
  ylab = "Frequency",
  main = "Histogram of Pitch Counts",
  axes = FALSE
)

axis(1, at = seq(0, 12, 2), labels = seq(0, 12, 2), col.axis = "#2d3436", 
     cex.axis = 1.1)
axis(2, at = seq(0, 230000, 20000), col.axis = "#2d3436", cex.axis = 1.1)

abline(h = seq(0, 230000, 20000), col = "#dfe6e9", lty = "dotted")
abline(v = seq(0, 12, 2), col = "#dfe6e9", lty = "dotted")

text(
  x = h$mids[h$counts > 0],
  y = h$counts[h$counts > 0] + 6000,
  labels = h$counts[h$counts > 0],
  col = "#0984e3",
  font = 2,
  cex = .5
)

box(col = "#636e72", lwd = 1.2)
```
:::

# Home Run Breakdowns   {background="#6B6B6B"}

## HR by Handedness  {background="#a6bee3"}
- Foul advantage boosted by favorable handedness
- In unfavorable matchup, high foul count still beneficial

::: {.center}
```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 4.5

query9 <- "WITH foul_balls AS (
  SELECT *,
         LENGTH(pitches) - LENGTH(REPLACE(pitches, 'F', '')) AS foul_ball_count,
         CASE 
           WHEN bathand = 'R' AND pithand = 'R' THEN 'RHH vs RHP'
           WHEN bathand = 'L' AND pithand = 'L' THEN 'LHH vs LHP'
           WHEN bathand = 'R' AND pithand = 'L' THEN 'RHH vs LHP'
           WHEN bathand = 'L' AND pithand = 'R' THEN 'LHH vs RHP'
           ELSE NULL
         END AS handedness_matchup
  FROM stat405baseball
),
filtered AS (
  SELECT * FROM foul_balls WHERE handedness_matchup IS NOT NULL
),
grouped AS (
  SELECT foul_ball_count, handedness_matchup, COUNT(*) AS total_at_bats,
         AVG(hr * 1.0) AS hr_rate
  FROM filtered
  GROUP BY foul_ball_count, handedness_matchup
),
filtered_min10 AS (
  SELECT * FROM grouped WHERE total_at_bats >= 10
),
sum_hr_rate AS (
  SELECT foul_ball_count, SUM(hr_rate) AS total_hr_rate
  FROM filtered_min10
  GROUP BY foul_ball_count
)
SELECT f.foul_ball_count, f.handedness_matchup, f.total_at_bats, f.hr_rate,
       f.hr_rate / s.total_hr_rate AS prop_hr_rate
FROM filtered_min10 f
JOIN sum_hr_rate s ON f.foul_ball_count = s.foul_ball_count
ORDER BY f.foul_ball_count, f.handedness_matchup;"
library(ggplot2)
library(scales)
library(viridis)

hrmatchup <- dbGetQuery(dcon, query9)

ggplot(hrmatchup, aes(x = foul_ball_count, y = prop_hr_rate, 
                      fill = handedness_matchup)) +
  geom_area(alpha = 0.7, color = "white", linewidth = 0.3) +
  scale_fill_viridis_d(name = "Matchup", option = "D") +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(
    title = "HR Rate by Foul Ball Count and Matchup",
    x = "Foul Balls",
    y = "Home Run Rate (Proportion)"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 12),
    axis.title = element_text(face = "bold"),
    axis.text = element_text(size = 11),
    panel.grid.major = element_line(color = "grey85"),
    panel.grid.minor = element_blank(),
    legend.position = "top",
    legend.title = element_text(face = "bold")
  )

```
:::

## HR by Foul Counts  {background="#a6bee3"}
- HR rate increases in AB with 9+ fouls
- Suggests that longer AB is more beneficial for batters

::: {.center}
```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 4.5
query3 <- "WITH foul_counts AS (
  SELECT *,
         LENGTH(pitches) - LENGTH(REPLACE(pitches, 'F', '')) AS foul_ball_count
  FROM stat405baseball
),
aggregated AS (
  SELECT foul_ball_count,
         COUNT(*) AS total_at_bats,
         AVG(hr) AS hr_rate
  FROM foul_counts
  GROUP BY foul_ball_count
)
SELECT *
FROM aggregated
WHERE total_at_bats >= 10
ORDER BY foul_ball_count;"
filtered_data <- dbGetQuery(dcon, query3)

ggplot(filtered_data, aes(x = foul_ball_count, y = hr_rate)) +
  geom_segment(aes(xend = foul_ball_count, y = 0, yend = hr_rate),
               color = "#4682B4", linewidth = 1.2, alpha = 0.7) +
  geom_point(color = "#D73027", size = 5) +
  geom_text(aes(label = round(hr_rate, 2)), vjust = -1, size = 4, 
            color = "black") +
  labs(
    title = "Home Run Rate by Number of Foul Balls in AB",
    x = "Number of Foul Balls",
    y = "Home Run Rate"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 12),
    axis.title = element_text(face = "bold"),
    panel.grid.major = element_line(color = "grey80"),
    panel.grid.minor = element_blank()
  )
```
:::

# Results by Ball/Strike Count   {background="#6B6B6B"}

## SLG by Count  {background="#a6bee3"}

- 3-ball counts or 2-0 result in best SLG
- Pitcher's counts have lower SLG, as expected

::: {.center}
```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 4.5
base_values <- c(single = 1, double = 2, triple = 3, hr = 4, walk = 1, iw = 1, xi = 1)
data <- dbGetQuery(dcon, "SELECT * FROM stat405baseball")
slugging_data <- data %>%
  filter(!(count %in% c(40, 41, 42, 43, 3, 13, 23, 33))) %>%
  mutate(
    total_bases = single * base_values["single"] +
      double * base_values["double"] +
      triple * base_values["triple"] +
      hr * base_values["hr"] +
      walk * base_values["walk"] +
      iw * base_values["iw"] +
      xi * base_values["xi"],
    total_at_bats = (single + double + triple + hr + walk + iw + xi + othout)  # Assuming 1 for out as part of PA
  ) %>%
  group_by(count) %>%
  summarise(
    total_bases = sum(total_bases),
    total_at_bats = sum(total_at_bats),
    slugging_pct = total_bases / total_at_bats
  ) %>%
  mutate(slugging_diff = slugging_pct - mean(slugging_pct))  # Difference from the league average slugging

ggplot(slugging_data, aes(x = reorder(count, slugging_diff), y = slugging_diff, 
                          fill = slugging_diff > 0)) +
  geom_bar(stat = "identity", width = 0.7, color = "gray20") +
  coord_flip() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40", 
             linewidth = 0.8) +
  scale_fill_manual(
    values = c("red", "steelblue"),
    labels = c("Below League Avg", "Above League Avg"),
    name = "SLG Diff"
  ) +
  labs(
    title = "Slugging Percentage by Count Against League Average",
    x = "Count",
    y = "SLG Difference"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 12),
    axis.title = element_text(face = "bold"),
    axis.text = element_text(size = 12),
    legend.position = "top"
  )
```
:::

## OBP by Count  {background="#a6bee3"}

- Only 3-ball counts have above league avg OBP
- Expected this result because OBP includes BB/HBP

::: {.center}
```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 4.5

library(dplyr)
library(ggplot2)
library(DBI)
library(RSQLite)
query7 <- "WITH filtered AS (
  SELECT *
  FROM stat405baseball
  WHERE count NOT IN (40, 41, 42, 43, 3, 13, 23, 33)
),
with_obp AS (
  SELECT
    count,
    (single + double + triple + hr + walk + iw + xi) AS reached_base,
    (single + double + triple + hr + walk + iw + xi + othout) AS total_pa
  FROM filtered
),
grouped AS (
  SELECT
    count,
    SUM(reached_base) AS total_reached_base,
    SUM(total_pa) AS total_pa,
    SUM(reached_base) * 1.0 / SUM(total_pa) AS obp
  FROM with_obp
  GROUP BY count
),
with_league_avg AS (
  SELECT 
    *,
    AVG(obp) OVER () AS league_obp
  FROM grouped
)
SELECT 
  count,
  total_reached_base,
  total_pa,
  obp,
  obp - league_obp AS obp_diff
FROM with_league_avg
ORDER BY obp_diff DESC;
"

library(ggplot2)

obp_data <- dbGetQuery(dcon, query7)
obp_data$count <- factor(obp_data$count)

ggplot(obp_data, aes(x = reorder(count, obp_diff), y = obp_diff, 
                     color = obp_diff > 0)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50", 
             linewidth = 0.8) +
  geom_segment(aes(xend = count, y = 0, yend = obp_diff), linewidth = 1.2) +  
  geom_point(size = 5) + 
  scale_color_manual(
    values = c("red", "steelblue"),
    labels = c("Below Avg", "Above Avg"),
    name = "OBP Diff"
  ) +
  labs(
    title = "On-Base Percentage by Count vs League Average",
    subtitle = "Positive values = better than league average",
    x = "Count",
    y = "Difference from League-Wide OBP"
  ) +
  coord_flip() +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 12),
    axis.title = element_text(face = "bold"),
    axis.text = element_text(size = 12),
    legend.position = "top"
  )
```
:::

## Killer  {background="#a6bee3"}

::: {.center}
```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 4.5
library(ggplot2)
library(dplyr)
library(grid)
library(shiny)
library(shinyjs)

# Example dataset for the sake of demonstration
killerdata <- data.frame(
  last_first = c("Batter A", "Batter B", "Batter C", "Batter D"),
  exit_velocity_avg = c(90, 92, 88, 85),
  xslg = c(0.450, 0.480, 0.400, 0.410),
  bb_percent = c(0.12, 0.10, 0.15, 0.13),
  whiff_percent = c(0.20, 0.18, 0.25, 0.22),
  ppa = c(3.8, 4.0, 3.5, 3.7)
)

# Example batter score weights
batter_score_weights <- c(
  xslg = 0.4,
  bb_percent = 0.3,
  whiff_percent = 0.2,
  exit_velocity_avg = 0.1
)

# Define percentile labels for the stats
stat_labels <- c(
  xslg = "Expected SLG",
  bb_percent = "Walk Rate (%)",
  whiff_percent = "Whiff Rate (%)",
  exit_velocity_avg = "Exit Velocity (mph)"
)

# Introduction slide content
cat("## Introduction\n\nThis presentation covers an in-depth analysis of the factors that influence batting performance in baseball.\n\nWe will explore various metrics that assess the quality of at-bats, including walk rate, strikeouts, exit velocity, and more.")

# Data overview slide content
cat("## Data Overview\n\nThe dataset consists of several metrics for individual batters including:\n\n- Walk rate (bb_percent)\n- Expected SLG (xslg)\n- Whiff rate (whiff_percent)\n- Pitches per plate appearance (ppa)\n- Exit velocity average (exit_velocity_avg)\n\nLet's start by exploring the first few rows of the dataset.")

# Display first few rows of data
head(killerdata)

# Metrics and visualizations slide content
cat("## Metrics and Visualizations\n\nWe will now visualize some key metrics that describe batter performance.\n\nThe following plot shows the relationship between exit velocity and expected SLG (xslg).")

# Plot Exit Velocity vs Expected SLG
ggplot(killerdata, aes(x = exit_velocity_avg, y = xslg)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "Exit Velocity vs Expected SLG", x = "Exit Velocity (mph)", y = "Expected SLG") +
  theme_minimal()

# Batter score calculation slide content
cat("## Batter Score Calculation\n\nWe use a weighted scoring method to calculate a composite score for each batter based on key metrics.\n\nThe weighted factors include:\n- Expected SLG (xslg)\n- Walk rate (bb_percent)\n- Whiff rate (whiff_percent)\n- Average exit velocity (exit_velocity_avg)\n\nNext, let's visualize the distribution of scores across batters.")

# Calculate and plot the distribution of batter scores
killerdata$batter_score <- rowSums(killerdata[, names(batter_score_weights)] * batter_score_weights, na.rm = TRUE)

ggplot(killerdata, aes(x = batter_score)) +
  geom_histogram(binwidth = 1, fill = "blue", color = "black") +
  labs(title = "Distribution of Batter Scores", x = "Batter Score", y = "Frequency") +
  theme_minimal()

# Shiny App for Batter Comparison
fluidPage(
  titlePanel("Batter Comparison Killer Plot"),
  sidebarLayout(
    sidebarPanel(
      selectInput("batter1", "Select Batter 1:", choices = unique(killerdata$last_first)),
      selectInput("batter2", "Select Batter 2:", choices = unique(killerdata$last_first),
                  selected = unique(killerdata$last_first)[2])
    ),
    mainPanel(
      splitLayout(cellWidths = c("50%", "50%"),
                  plotOutput("plot1", height = "700px"),
                  plotOutput("plot2", height = "700px")),
      br(),
      div(
        style = "text-align: center; margin-top: -10px;",
        span(textOutput("preferredBatter"),
             style = "font-size: 26px; font-weight: bold; color: black;"),
        br(),
        span(textOutput("scoreDifference"),
             style = "font-size: 18px; font-weight: normal; color: gray;")
      )
    )
  )
)

# Shiny server function
server <- function(input, output) {

  get_player_data <- reactive({
    req(input$batter1, input$batter2)
    killerdata %>%
      filter(last_first %in% c(input$batter1, input$batter2))
  })

  normalize_stat <- function(x) {
    ecdf(x)(x) * 100
  }

  draw_profile_card <- function(player_row, player_percentiles, opponent_percentiles, score_color) {
    stats <- names(player_percentiles)
    stat_labels_clean <- stat_labels[stats]

    player_score <- sum(player_percentiles[stats] * batter_score_weights[stats], na.rm = TRUE)

    grid.newpage()
    pushViewport(viewport(layout = grid.layout(length(stats) + 3, 3,
                                               widths = unit(c(0.4, 0.4, 0.2), "npc"))))

    # Player name
    grid.text(player_row$last_first,
              vp = viewport(layout.pos.row = 1, layout.pos.col = 1:2),
              gp = gpar(fontsize = 20, fontface = "bold"))

    # Percentile header
    grid.text("Percentile",
              vp = viewport(layout.pos.row = 1, layout.pos.col = 3),
              gp = gpar(fontsize = 12, fontface = "bold"))

    for (i in seq_along(stats)) {
      stat <- stats[i]
      p_val <- player_percentiles[stat]
      o_val <- opponent_percentiles[stat]

      p_val_clean <- ifelse(is.na(p_val), 0, p_val)
      o_val_clean <- ifelse(is.na(o_val), 0, o_val)

      # Cap values to range [0, 100]
      p_val_clean <- max(min(p_val_clean, 100), 0)
      o_val_clean <- max(min(o_val_clean, 100), 0)

      # Convert to position on bar between 0.05 and 0.95 (to stay inside)
      pct_to_pos <- function(pct) {
        0.05 + 0.9 * (pct / 100)
      }

      bg_color <- if (p_val_clean > o_val_clean) {
        "palegreen3"
      } else if (p_val_clean < o_val_clean) {
        "indianred2"
      } else {
        "gray80"
      }

      row_index <- i + 1

      for (col in 1:3) {
        grid.rect(gp = gpar(col = NA, fill = bg_color),
                  vp = viewport(layout.pos.row = row_index, layout.pos.col = col))
      }

      # Stat label
      grid.text(stat_labels_clean[i],
                just = "left", x = unit(0.01, "npc"),
                gp = gpar(fontsize = 10, fontface = "bold"),
                vp = viewport(layout.pos.row = row_index, layout.pos.col = 1))

      # Percentile bar background
      grid.rect(x = unit(0.5, "npc"), width = unit(0.9, "npc"), height = unit(0.6, "lines"),
                gp = gpar(col = NA, fill = "gray90"),
                vp = viewport(layout.pos.row = row_index, layout.pos.col = 2))

      # Tick mark
      grid.rect(x = unit(pct_to_pos(p_val_clean), "npc"),
                width = unit(0.015, "npc"), height = unit(0.6, "lines"),
                gp = gpar(col = NA, fill = "black"),
                vp = viewport(layout.pos.row = row_index, layout.pos.col = 2))

      # Bold percentile value
      grid.text(round(p_val_clean),
                x = unit(0.5, "npc"), y = unit(0.5, "npc"),
                gp = gpar(fontsize = 9, fontface = "bold"),
                vp = viewport(layout.pos.row = row_index, layout.pos.col = 3))
    }

    # Batter Score (bold and colored)
    grid.text(paste("Score:", round(player_score, 1)),
              gp = gpar(fontsize = 14, fontface = "bold", col = score_color),
              vp = viewport(layout.pos.row = length(stats) + 2, layout.pos.col = 1:3))
  }

  calculate_scores <- reactive({
    stats <- names(stat_labels)
    percentiles_df <- killerdata
    for (stat in stats) {
      percentiles_df[[paste0(stat, "_pct")]] <- normalize_stat(percentiles_df[[stat]])
    }

    p1 <- killerdata %>% filter(last_first == input$batter1)
    p2 <- killerdata %>% filter(last_first == input$batter2)

    p1_percentiles <- percentiles_df %>%
      filter(last_first == input$batter1) %>%
      select(paste0(stats, "_pct"))
    p2_percentiles <- percentiles_df %>%
      filter(last_first == input$batter2) %>%
      select(paste0(stats, "_pct"))

    list(p1_percentiles = p1_percentiles, p2_percentiles = p2_percentiles)
  })

  output$plot1 <- renderPlot({
    scores <- calculate_scores()
    draw_profile_card(scores$p1_percentiles, scores$p2_percentiles, "blue")
  })

  output$plot2 <- renderPlot({
    scores <- calculate_scores()
    draw_profile_card(scores$p2_percentiles, scores$p1_percentiles, "red")
  })

  output$preferredBatter <- renderText({
    scores <- calculate_scores()
    p1_score <- sum(scores$p1_percentiles * batter_score_weights)
    p2_score <- sum(scores$p2_percentiles * batter_score_weights)

    if (p1_score > p2_score) {
      paste(input$batter1, "is the better batter!")
    } else if (p1_score < p2_score) {
      paste(input$batter2, "is the better batter!")
    } else {
      "Both batters are equally good!"
    }
  })

  output$scoreDifference <- renderText({
    scores <- calculate_scores()
    p1_score <- sum(scores$p1_percentiles * batter_score_weights)
    p2_score <- sum(scores$p2_percentiles * batter_score_weights)
    paste("Score difference: ", round(abs(p1_score - p2_score), 2))
  })
}
```
:::